---
title: "QTL mapping of hypertension in mice"
subtitle: "R/qtl exercise using the hyper cross"
format: html
editor: visual
---

## 1. Install and load packages

```{r}
#| label: setup

# renv::install("qtl")

library(qtl)
```

## 2. Load data

```{r}
data(hyper, package = "qtl")
```

## 3. Inspect the cross

```{r}
#| label: cross-summary

summary(hyper)

nind(hyper)
nphe(hyper)
nchr(hyper)
totmar(hyper)
nmar(hyper)
```

**Observations**

The `hyper` dataset is a backcross of 250 mice with two phenotypes, both recorded for all individuals. It contains 174 markers across 19 autosomes and the X chromosome. Marker coverage varies by chromosome, and only 47.7% of the possible genotype observations are present. The autosomal genotype classes are nearly balanced (BB: 50.1%; BA: 49.9%). The extent and pattern of missing genotype data warrant closer inspection in the plots below.

```{r}
#| label: Plotting helper function

save_plot <- function(filename, draw, width = 10, height = 7) {
  pdf(
    file = file.path("results", "plots", filename),
    width = width,
    height = height
  )
  on.exit(dev.off(), add = TRUE)
  draw()
}
```

**Overview of missing genotypes**

```{r}
#| label: fig-missing
#| fig-cap: "Missing genotype observations across individuals and markers."
#| fig-width: 13
#| fig-height: 7

old_par <- par(
  mar = c(5, 4, 7, 2) + 0.1,
  cex.axis = 0.85,
  cex.main = 1.3,
  cex.lab = 1.25  
)

plotMissing(
  hyper,
  alternate.chrid = TRUE,
  main = "Missing genotypes across individuals and markers"
)

par(old_par)
```

```{r}
#| label: marker-coverage-hyper

marker_coverage <- do.call(
  rbind,
  lapply(names(hyper$geno), function(chr) {
    genotypes <- hyper$geno[[chr]]$data
    
    data.frame(
      chromosome = chr,
      marker = colnames(genotypes),
      genotyped = colSums(!is.na(genotypes)),
      missing = colSums(is.na(genotypes)),
      row.names = NULL
    )
  })
)

# The marker(s) with no genotype data
marker_coverage[marker_coverage$genotyped == 0, ]

# Markers typed in at most 100 of the 250 mice
marker_coverage[marker_coverage$genotyped <= 100, ]
```

```{r}
print(marker_coverage[marker_coverage$genotyped <= 100, ])
```

**Observations**

Genotype coverage is strongly structured. Numerous markers across the genome have observations for exactly 92 of the 250 mice and are missing for the other 158, producing the broad horizontal boundary in the plot. Coverage is even lower at selected markers: `D15Mit175` on chromosome 15 is typed in 13 mice, `D5Mit61` on chromosome 5 in 15, and `D4Mit53` and `D4Mit89` on chromosome 4 in 20 each. The chromosome 14 marker `D14Mit48` has no observed genotypes at all (0/250). Thus, missingness varies both among groups of mice and among individual markers; `D14Mit48` can be removed before subsequent analyses.

**Genetic map across**

```{r}
#| label: fig-map
#| fig-cap: "Original genetic map across the 19 autosomes and X chromosome."

plotMap(
  hyper,
  main = "Genetic map of the hyper cross"    
)
```

**Observations**

The original genetic map places 174 markers across chromosomes 1–19 and X. Each horizontal tick represents a marker position in centimorgans (cM). Chromosome 1 has the longest plotted span, approximately 3–116 cM, while chromosome 18 spans only about 0–35 cM. Marker spacing is uneven: several markers cluster around 15–35 cM on chromosome 4 and near 0–10 cM on chromosome 17, whereas other intervals have relatively few markers. This plot shows their mapped positions, but it does not show how many mice were successfully genotyped at each marker.

**Distribution of blood pressure in the 250 mice**

```{r}
#| label: fig-bp
#| fig-cap: "Distribution of blood pressure in the 250 mice."
#| fig-width: 8
#| fig-height: 5

plotPheno(
  hyper,
  pheno.col = "bp",
  main = "Blood pressure distribution",
  xlab = "Blood pressure"
)
```

**Observations**

Blood pressure varies continuously across the 250 mice, from approximately 82 to 128 in the plotted values. Most observations fall between about 90 and 110, with the highest frequencies around 94–100 and 104–106. Only a few mice have values below 88 or above 120. The distribution has no clear separation into distinct groups, making blood pressure suitable for the quantitative QTL analysis that follows.

**Distribution of mice by sex**

```{r}
#| label: fig-sex

#| fig-cap: "Distribution of mice by sex."
#| fig-width: 7
#| fig-height: 5

plotPheno(
  hyper,
  pheno.col = "sex",
  main = "Distribution of sex in the hyper cross",
  xlab = "Sex"
)
```

**Observations**

All 250 individuals in the `hyper` cross are recorded as male; none are recorded as female. Sex therefore does not vary in this dataset, so it cannot be used to compare blood pressure between males and females.

**Marker names on selected chromosomes**

```{r}
#| label: fig-selected-markers
#| fig-cap: "Marker names on selected chromosomes."
#| fig-width: 12
#| fig-height: 9

plotMap(
  hyper,
  chr = c(1, 4, 6, 7, 15),
  show.marker.names = TRUE
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 1."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 1,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 1"
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 4."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 4,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 4"
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 6."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 6,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 6"
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 7."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 7,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 7"
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 15."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 15,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 15"
)
```

```{r}
#| label: marker-on-chromosme-1-chromosme-15

hyper$geno[["1"]]$map
hyper$geno[["15"]]$map
```

**Observations**

On chromosome 1, D1Mit46 and D1Mit132 both map to 43.7 cM, while D1Mit14, D1Mit105, D1Mit159, and D1Mit267 all map to 82.0 cM. On chromosome 15, D15Mit11 and D15Mit175 both map to 5.5 cM; D15Mit56 and D15Mit22 map to 16.4 cM; and D15Mit206 and D15Mit152 map to 17.5 cM. Their labels overlap because they share map positions, rather than because the plots are too narrow.

**Remove the marker without genotype data**

```{r}
#| label: remove-null-markers

markers_before <- totmar(hyper)

hyper <- drop.nullmarkers(hyper)

markers_after <- totmar(hyper)

data.frame(
  stage = c("Before", "After"),
  markers = c(markers_before, markers_after)
)
```

## 2. Recombination fractions and the genetic map

```{r}
#| label: estimate-recombination

hyper_est <- est.rf(hyper)
```

**Recombination pattern on all chromosomes**

```{r}
#| label: fig-recombination-all
#| fig-cap: "Pairwise recombination fractions and associated LOD scores."
#| fig-width: 11
#| fig-height: 10

plotRF(hyper_est)
```
