---
title: "QTL mapping of hypertension in mice"
subtitle: "R/qtl exercise using the hyper cross"
format: html
editor: visual
---

## 1. Install and load packages

```{r}
#| label: setup

# renv::install("qtl")

library(qtl)
```

## 2. Load data

```{r}
data(hyper, package = "qtl")
```

## 3. Inspect the cross

```{r}
#| label: cross-summary

summary(hyper)

nind(hyper)
nphe(hyper)
nchr(hyper)
totmar(hyper)
nmar(hyper)
```

**Observations**

The `hyper` dataset is a backcross of 250 mice with two phenotypes, both recorded for all individuals. It contains 174 markers across 19 autosomes and the X chromosome. Marker coverage varies by chromosome, and only 47.7% of the possible genotype observations are present. The autosomal genotype classes are nearly balanced (BB: 50.1%; BA: 49.9%). The extent and pattern of missing genotype data warrant closer inspection in the plots below.

```{r}
#| label: Plotting helper function

save_plot <- function(filename, draw, width = 10, height = 7) {
  pdf(
    file = file.path("results", "plots", filename),
    width = width,
    height = height
  )
  on.exit(dev.off(), add = TRUE)
  draw()
}
```

**Overview of missing genotypes**

```{r}
#| label: fig-missing
#| fig-cap: "Missing genotype observations across individuals and markers."
#| fig-width: 13
#| fig-height: 7

old_par <- par(
  mar = c(5, 4, 7, 2) + 0.1,
  cex.axis = 0.85,
  cex.main = 1.3,
  cex.lab = 1.25  
)

plotMissing(
  hyper,
  alternate.chrid = TRUE,
  main = "Missing genotypes across individuals and markers"
)

par(old_par)
```

```{r}
#| label: marker-coverage-hyper

marker_coverage <- do.call(
  rbind,
  lapply(names(hyper$geno), function(chr) {
    genotypes <- hyper$geno[[chr]]$data
    
    data.frame(
      chromosome = chr,
      marker = colnames(genotypes),
      genotyped = colSums(!is.na(genotypes)),
      missing = colSums(is.na(genotypes)),
      row.names = NULL
    )
  })
)

# The marker(s) with no genotype data
marker_coverage[marker_coverage$genotyped == 0, ]

# Markers typed in at most 100 of the 250 mice
marker_coverage[marker_coverage$genotyped <= 100, ]
```

```{r}
print(marker_coverage[marker_coverage$genotyped <= 100, ])
```

**Observations**

Genotype coverage is strongly structured. Numerous markers across the genome have observations for exactly 92 of the 250 mice and are missing for the other 158, producing the broad horizontal boundary in the plot. Coverage is even lower at selected markers: `D15Mit175` on chromosome 15 is typed in 13 mice, `D5Mit61` on chromosome 5 in 15, and `D4Mit53` and `D4Mit89` on chromosome 4 in 20 each. The chromosome 14 marker `D14Mit48` has no observed genotypes at all (0/250). Thus, missingness varies both among groups of mice and among individual markers; `D14Mit48` can be removed before subsequent analyses.

**Genetic map across**

```{r}
#| label: fig-map
#| fig-cap: "Original genetic map across the 19 autosomes and X chromosome."

plotMap(
  hyper,
  main = "Genetic map of the hyper cross"    
)
```

**Observations**

The original genetic map places 174 markers across chromosomes 1–19 and X. Each horizontal tick represents a marker position in centimorgans (cM). Chromosome 1 has the longest plotted span, approximately 3–116 cM, while chromosome 18 spans only about 0–35 cM. Marker spacing is uneven: several markers cluster around 15–35 cM on chromosome 4 and near 0–10 cM on chromosome 17, whereas other intervals have relatively few markers. This plot shows their mapped positions, but it does not show how many mice were successfully genotyped at each marker.

**Distribution of blood pressure in the 250 mice**

```{r}
#| label: fig-bp
#| fig-cap: "Distribution of blood pressure in the 250 mice."
#| fig-width: 8
#| fig-height: 5

plotPheno(
  hyper,
  pheno.col = "bp",
  main = "Blood pressure distribution",
  xlab = "Blood pressure"
)
```

**Observations**

Blood pressure varies continuously across the 250 mice, from approximately 82 to 128 in the plotted values. Most observations fall between about 90 and 110, with the highest frequencies around 94–100 and 104–106. Only a few mice have values below 88 or above 120. The distribution has no clear separation into distinct groups, making blood pressure suitable for the quantitative QTL analysis that follows.

**Distribution of mice by sex**

```{r}
#| label: fig-sex

#| fig-cap: "Distribution of mice by sex."
#| fig-width: 7
#| fig-height: 5

plotPheno(
  hyper,
  pheno.col = "sex",
  main = "Distribution of sex in the hyper cross",
  xlab = "Sex"
)
```

**Observations**

All 250 individuals in the `hyper` cross are recorded as male; none are recorded as female. Sex therefore does not vary in this dataset, so it cannot be used to compare blood pressure between males and females.

**Marker names on selected chromosomes**

```{r}
#| label: fig-selected-markers
#| fig-cap: "Marker names on selected chromosomes."
#| fig-width: 12
#| fig-height: 9

plotMap(
  hyper,
  chr = c(1, 4, 6, 7, 15),
  show.marker.names = TRUE
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 1."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 1,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 1"
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 4."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 4,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 4"
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 6."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 6,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 6"
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 7."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 7,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 7"
)
```

```{r}
#| label: fig-chr4-markers
#| fig-cap: "Marker positions on chromosome 15."
#| fig-width: 8
#| fig-height: 12

plotMap(
  hyper,
  chr = 15,
  show.marker.names = TRUE,
  main = "Genetic map of chromosome 15"
)
```

```{r}
#| label: marker-on-chromosme-1-chromosme-15

hyper$geno[["1"]]$map
hyper$geno[["15"]]$map
```

**Observations**

On chromosome 1, D1Mit46 and D1Mit132 both map to 43.7 cM, while D1Mit14, D1Mit105, D1Mit159, and D1Mit267 all map to 82.0 cM. On chromosome 15, D15Mit11 and D15Mit175 both map to 5.5 cM; D15Mit56 and D15Mit22 map to 16.4 cM; and D15Mit206 and D15Mit152 map to 17.5 cM. Their labels overlap because they share map positions, rather than because the plots are too narrow.

**Remove the marker without genotype data**

```{r}
#| label: remove-null-markers

markers_before <- totmar(hyper)

hyper <- drop.nullmarkers(hyper)

markers_after <- totmar(hyper)

data.frame(
  stage = c("Before", "After"),
  markers = c(markers_before, markers_after)
)
```

## 4. Recombination fractions and the genetic map

```{r}
#| label: estimate-recombination

hyper_est <- est.rf(hyper)
```

**Recombination pattern on all chromosomes (Whole-genome recombination)**

```{r}
#| label: fig-recombination-all
#| fig-cap: "Pairwise recombination fractions and associated LOD scores."
#| fig-width: 11
#| fig-height: 10

plotRF(hyper_est, cex = 1.5)
```

**Observations**

The clear blocks along the diagonal show stronger linkage between markers on the same chromosome. Most comparisons between chromosomes show weaker linkage. The irregular stripes within some chromosome blocks indicate that the pairwise estimates are not equally consistent throughout the genome.

**Recombination pattern on chromosomes 1 and 4**

```{r, fig.width=9, fig.height=9}
#| label: fig-recombination-1-4

#| fig-cap: "Pairwise recombination fractions and linkage LOD scores for chromosomes 1 and 4."
#| echo: false
#| out-width: "100%"
#| fig-align: center

old_par <- par(
  ann = FALSE,
  mar = c(5, 5, 5, 5) + 0.1,
  pty = "s"
)

plotRF(
  hyper_est,
  chr = c(1, 4),
  what = "both",
  main = ""
)

mtext(
  "Pairwise linkage: chromosomes 1 and 4",
  side = 3, line = 3,
  font = 2, cex = 1.5
)

mtext("Marker index", side = 1, line = 3, cex = 1.3)
mtext("Marker index", side = 2, line = 3, cex = 1.3)

par(old_par)
```

**Observations**

The two chromosomes form distinct linkage blocks. Within each chromosome, nearby markers generally show lower recombination fractions and stronger linkage than more distant markers. Comparisons between chromosomes 1 and 4 mostly show weak linkage. The patchy colors and gray cells indicate uneven information for some marker pairs; gray cells represent unavailable estimates, rather than evidence of absent linkage.

**Recombination patterns on chromosome 6**

```{r}
#| label: fig-recombination-6
#| fig-cap: "Pairwise recombination fractions and linkage LOD scores for markers on chromosome 6."

plotRF(
  hyper_est,
  chr = 6,
  what = "both",
  main = "Pairwise linkage between markers on chromosome 6"
)
```

**Observations**

Marker 10 (D6Mit201) shows unusually high estimated recombination fractions with several chromosome 6 markers, including nearby markers 8 and 11, visible as dark purple cells above the diagonal. This interrupts the brighter pattern generally seen between neighboring markers. The worksheet attributes this unusual pattern to selective genotyping: marker 10 was typed primarily in individuals showing recombination in the surrounding interval, which can inflate its estimated recombination fractions.

**Missing genotype pattern on chromosome 6**

```{r}
#| label: fig-missing-6
#| fig-cap: "Missing genotype pattern on chromosome 6."

plotMissing(
  hyper_est, 
  chr = 6,
  main = "Missing genotype pattern on chromosome 6"
)
```

**Observations**

Marker 10 was genotyped in only a small, scattered subset of individuals. Markers 2, 3, 5, 7, and 9 also have extensive missing data, whereas markers 1, 4, 6, 8, and 11 have much more complete coverage.

#### Estimate a new map

```{r}
#| label: estimate-map

newmap <- est.map(hyper_est, error.prob = 0.01)
```

```{r}
#| label: fig-map-comparison

#| fig-cap: "Original and newly estimated genetic maps."
#| fig-width: 12
#| fig-height: 9

plotMap(hyper_est, newmap)
```

**Observations**

Re-estimating the genetic map while keeping marker order fixed changes the distances between markers. The most pronounced expansion occurs on chromosomes 6, 13, and 18: their total lengths increase approximately from 85 to 120 cM, 70 to 115 cM, and 45 to 80 cM, respectively. Chromosome 11 also expands visibly, from approximately 75 to 115 cM. The connecting lines show that changes in spacing are uneven across each chromosome.

## 5. Identify possible genotyping errors

```{r}
#| label: error-lod

hyper_err_lod <- calc.errorlod(hyper_est, error.prob = 0.01)

top.errorlod(hyper_err_lod)
```

**Observations**

The analysis flagged 128 genotype calls across 22 markers on chromosomes 1, 11, 15, 16, and 17. The highest error LOD scores occurred at D11Mit79 in individuals 152 and 144 (13.36 and 13.18), followed by D1Mit14 in individuals 118, 159, 162, and 170 (13.11 each). Several marker pairs were flagged in the same individuals; notably, D16Mit106 and D16Mit70 were both flagged in individuals 16, 23, 29, 31, 38, 40, 45, and 87. These repeated patterns indicate genotype inconsistencies that require inspection alongside neighboring markers and their map positions. High error LOD scores identify possible errors, but do not confirm incorrect calls, because genuine recombination between very tightly linked markers can also produce high scores.

**Observe genotypes for selected individuals on chromosome 16**

```{r}
#| label: fig-genotypes-16

#| fig-cap: "Observed genotypes for selected individuals on chromosome 16."
#| fig-width: 12
#| fig-height: 8

plotGeno(hyper_err_lod, chr = 16, ind = c(24:34, 71:81))
```

**Observations**

White circles represent AA, black circles represent AB, blue crosses indicate genotype transitions consistent with crossovers, and red squares flag possible errors above the default error LOD cutoff of 4. Individuals 29 and 31 show conflicting genotypes at D16Mit106 and D16Mit70 (identified from the above error LOD table), the two closely spaced markers near 52 cM. Both marker calls are flagged by red squares, consistent with the high error LOD scores reported in the table. These individuals have predominantly AB genotypes across the preceding markers, followed by an AA call and a return to AB at the chromosome end. This pattern implies two genotype transitions within a short interval and therefore raises suspicion of a genotyping error. However, the plot alone cannot establish which call is incorrect; genuine recombination or inaccurate marker spacing could also contribute. The other displayed individuals have no calls flagged above the cutoff.

## 6. Single-QTL genome scans

### 6.1 Calculate genotype probabilities

```{r}
#| label: genotype-probabilities

hyper_genoprob <- calc.genoprob(
  hyper_err_lod,
  step = 1,
  error.prob = 0.01
)
```

### 6.2 EM and Haley–Knott scans

```{r}
#| label: em-hk-scans

out.em <- scanone(hyper_genoprob, pheno.col = "bp", method = "em")
out.hk <- scanone(hyper_genoprob, pheno.col = "bp", method = "hk")
```

### 6.3 Multiple-imputation scan

```{r}
#| label: imputation-scan

set.seed(20260929)

hyper_imput <- sim.geno(
  hyper_genoprob,
  step = 2,
  n.draws = 16,
  error.prob = 0.01
)

out.imp <- scanone(hyper_imput, pheno.col = "bp", method = "imp")
```

### 6.4 Compare scan results

```{r}
#| label: scan-summaries

summary(out.em)
summary(out.em, threshold = 3)
summary(out.hk, threshold = 3)
summary(out.imp, threshold = 3)
```

```{r}
#| label: highest-scan-peaks

max(out.em)
max(out.hk)
max(out.imp)
```

```{r}
#| label: fig-em-scan
#| fig-cap: "EM single-QTL scan for blood pressure on chromosomes 1, 4, and 15."
#| fig-width: 10
#| fig-height: 6

plot(
  out.em,
  chr = c(1, 4, 15),
  main = "Blood pressure: EM single-QTL scan",
  xlab = "Chromosome",
  ylab = "LOD score",
  col = "black",
  lwd = 2,
  cex.main = 1.5,
  cex.lab = 1.2,
  cex.axis = 1.1
)
```

```{r}
#| label: fig-three-scans
#| fig-cap: "Comparison of EM, Haley–Knott, and multiple-imputation single-QTL scans for blood pressure."
#| fig-width: 10
#| fig-height: 6

scan_colors <- c(
  EM = "#0072B2",
  HK = "#D55E00",
  Imputation = "#009E73"
)

plot(
  out.em,
  out.hk,
  out.imp,
  chr = c(1, 4, 15),
  main = "Blood pressure: comparison of single-QTL scans",
  xlab = "Chromosome",
  ylab = "LOD score",
  col = unname(scan_colors),
  lty = c(1, 2, 3),
  lwd = 2,
  cex.main = 1.5,
  cex.lab = 1.2,
  cex.axis = 1.1
)

legend(
  "topright",
  legend = c("EM", "Haley–Knott", "Multiple imputation"),
  col = unname(scan_colors),
  lty = c(1, 2, 3),
  lwd = 2,
  bty = "n",
  cex = 1.1
)
```

**Observations and interpretation**

All three methods identify the strongest genome-wide peak on chromosome 4 at 29.5 cM, corresponding to marker D4Mit164, with LOD scores of 8.02–8.09. Their agreement indicates that this candidate QTL signal is consistent across the three analysis methods. A weaker signal occurs on chromosome 1, with peak LOD scores of 3.49–3.55. EM and Haley–Knott place this peak at 48.3 cM, whereas multiple imputation places it at 49.2 cM, at marker D1Mit334.

The curves largely overlap, particularly around the chromosome 4 peak. Small differences may reflect the methods’ different treatment of genotype uncertainty, the 1 cM versus 2 cM calculation grids, and simulation variability from the 16 imputation draws. Chromosome 15 shows a weaker signal; its EM peak is LOD 1.77 at 19.5 cM, and none of the three methods reports a chromosome 15 peak above the reporting threshold of 3.

These results prioritize chromosomes 4 and 1 for further investigation. However, LOD \> 3 is only the reporting cutoff used here; genome-wide statistical significance must be assessed using the permutation thresholds in the next section.

## 7. Genome-wide significance

### 7.1 Run Haley–Knott permutation tests

```{r}
#| label: hk-permutations

set.seed(20260929)

operm.hk <- scanone(
  hyper_genoprob,
  pheno.col = "bp",
  method = "hk",
  n.perm = 1000
)
```

### 7.2 Estimate the genome-wide 5% LOD threshold

```{r}
#| label: permutation-threshold

summary(operm.hk, alpha = 0.05)
```

### 7.3 Identify significant peaks and report adjusted p-values

```{r}
#| label: adjusted-scan-results

summary(
  out.hk,
  perms = operm.hk,
  alpha = 0.05,
  pvalues = TRUE
)
```

**Observations and interpretation**

Based on 1,000 Haley–Knott permutations, the estimated genome-wide 5% significance threshold is LOD 2.65. Two chromosome peaks exceed this threshold: chromosome 1 at 48.3 cM (LOD 3.55, adjusted p = 0.005) and chromosome 4 at 29.5 cM, at marker D4Mit164 (LOD 8.09, adjusted p displayed as 0.000). Both provide statistically significant evidence of a blood-pressure QTL, with substantially stronger evidence on chromosome 4.

The displayed p-value of 0.000 reflects the finite permutation resolution and output rounding; it should not be interpreted as an exact probability of zero. No other chromosome peak meets the genome-wide 5% threshold.

The earlier LOD cutoff of 3 was a descriptive reporting filter. In contrast, LOD 2.65 is estimated from the permutation distribution of genome-wide maximum LOD scores and accounts for searching across the genome.

## 8. Fit and refine a two-QTL model

### 8.1 Define the initial QTL locations

The initial locations are chromosome 1 at 48.3 cM and chromosome 4 at 29.5 cM. If a requested position is absent from the calculation grid, the nearest available position is used.

```{r}
#| label: make-two-qtls

qc <- c(1, 4)
qp <- c(48.3, 29.5)

qtl_model <- makeqtl(
  hyper_imput,
  chr = qc,
  pos = qp,
  what = "draws"
)

qtl_model
```

**Observation**

The initial two-QTL object contains Q1 on chromosome 1 at 49.2 cM and Q2 on chromosome 4 at 29.5 cM, using 16 genotype imputations. The requested chromosome 1 position of 48.3 cM was mapped to the nearest available imputation-grid position, 49.2 cM. These returned positions will serve as the starting points for model fitting and refinement.

### 8.2 Fit the additive two-QTL model

The formula includes both QTL main effects without an interaction.

```{r}
#| label: fit-additive-model

additive_formula <- y ~ Q1 + Q2

out.fq <- fitqtl(
  hyper_imput,
  pheno.col = "bp",
  qtl = qtl_model,
  formula = additive_formula,
  method = "imp"
)

summary(out.fq)
```

**Observations and interpretation**

The additive two-QTL model, fitted by multiple imputation using 250 individuals, has an overall LOD score of 12.67 and explains approximately 20.82% of the observed blood-pressure variance.

The drop-one-QTL analysis shows that both loci contribute after accounting for the other. Removing Q1 on chromosome 1 at 49.2 cM reduces the model LOD by 4.69, with a reported variance contribution of 7.14% and an F-test p = 3.97 × 10⁻⁶. Removing Q2 on chromosome 4 at 29.5 cM reduces the LOD by 9.18, with a variance contribution of 14.60% and p = 1.06 × 10⁻¹⁰. Thus, chromosome 4 contributes more strongly to the joint model.

These tests assess each QTL conditional on the other at the selected positions; their p-values are distinct from the genome-wide permutation-adjusted p-values in Section 7. The model includes no interaction, so this analysis does not test whether the two QTL interact.

### 8.3 Estimate QTL effects

```{r}
#| label: fit-estimated-effects

out.fq.effects <- fitqtl(
  hyper_imput,
  pheno.col = "bp",
  qtl = qtl_model,
  formula = additive_formula,
  method = "imp",
  get.ests = TRUE
)

summary(out.fq.effects)
```

**Observations and interpretation**

After accounting for the other QTL, heterozygotes have an estimated mean blood pressure 4.50 mmHg lower than homozygotes at Q1 on chromosome 1 and 6.45 mmHg lower at Q2 on chromosome 4. Both effects are negative, with standard errors of approximately 0.95–0.96 mmHg. The chromosome 4 QTL has the larger estimated effect, consistent with its stronger contribution to the model in Section 8.2. Because the model contains no interaction, these genotype effects are assumed to add together. This backcross comparison does not allow additive and dominance effects to be estimated separately.
